---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Load libraries to be used later

```{r, message=F, warning=F}
library(dplyr) #reloading here to make group_by work in Knitr
library(ggplot2) #reloading here to make group_by work in Knitr
```

Load the data (i.e. read.csv())
Read in the activity data from previously downloaded CSV dataset

```{r load_data}
activity <- read.csv("activity.csv")
```
For this part of the assignment, we ignore the missing values in the dataset.

1. Calculate the total number of steps taken per day

```{r total_steps_perday}
totalsteps <- aggregate(steps ~ date, activity, sum)
```

2. Make a histogram of the total number of steps taken each day

```{r hist_total_steps_perday}
hist(totalsteps$steps, breaks=10,
     main = paste("Histogram of Total Steps per Day"), xlab = "Total Steps",
     ylab ="Frequency of Days")
```

3. Calculate and report the mean and median of the total number of steps taken per day

```{r meanmedian_total_steps_perday}
summary(totalsteps)
```

## What is mean total number of steps taken per day?

For this part of the assignment, we ignore the missing values in the dataset.

1. Calculate the total number of steps taken per day

```{r total_steps_perday}
totalsteps <- aggregate(steps ~ date, activity, sum)
```

2. Make a histogram of the total number of steps taken each day

```{r hist_total_steps_perday}
hist(totalsteps$steps, breaks=10,
     main = paste("Histogram of Total Steps per Day"), xlab = "Total Steps",
     ylab ="Frequency of Days")
```

3. Calculate and report the mean and median of the total number of steps taken per day

```{r meanmedian_total_steps_perday}
summary(totalsteps)
```


## What is the average daily activity pattern?

Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis)
and the average number of steps taken, averaged across all days (y-axis)

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

1. Find the mean of steps per interval and aggreate these into a new dataframe

```{r meansteps_interval}
stepinterval <- aggregate(steps ~ interval, activity, mean)
head(stepinterval)
```

2. Plot the data frame to show mean # of steps per 5 minute interval across all days of the dataset

```{r plot_meansteps_interval}

plot.new()
plot(stepinterval, type = "l", xlab ="Time of Day",
     ylab = "Mean number of steps",
     main = "Mean steps measured at 5 minute intervals", axes = FALSE)

axis(side=1, at = seq(0, 2400, by = 100))
axis(side=2, at = seq(0, 220, by = 10))
```

3. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

In the dataframe that's already been grouped by interval, find the row with the 
max number of steps, then display the row data.

```{r max_meansteps_interval}
stepinterval[which.max(stepinterval$steps), ]
```

## Imputing missing values

Note that there are a number of days/intervals where there are missing values
(coded as NA). The presence of missing days may introduce bias into some
calculations or summaries of the data.

1. Calculate and report the total number of missing values in the dataset (i.e.
the total number of rows with NAs)

```{r NA-count}
## Summary reports the number of NAs per variable
summary(activity)

## Or count them individually
sum(is.na(activity$steps))
sum(is.na(activity$interval))
sum(is.na(activity$date))

## Or create a dataframe with the count from individual columns
totalNAs <- sapply(activity, function(x) sum(is.na(x)))
totalNAs
```

### Filling in Missing values

Devise a strategy for filling in all of the missing values in the dataset. 

1. Create a new dataset that is equal to the original dataset but with the
missing data filled in.

```{r Fillin_missing_data}
## Duplicate original data to preserve it
activityNoNAs <- activity

## Add the means steps per interval across days data to the dataframe - since
## the intervals in activityNoNAs repeat on the same 288 interval cycle, simply
## adding the cycle to the records matches correctly.  This is a bit of a
## shortcut.

activityNoNAs$meansteps <- stepinterval$steps

## For each time the original steps data is NA, copy the meansteps data in to
## replace it.

head(activityNoNAs)

for (i in 1:nrow(activityNoNAs)){
if (is.na(activityNoNAs$steps[i])) {
  activityNoNAs$steps[i] <- activityNoNAs$meansteps[i]
}
}
```


2. Make a histogram of the total number of steps taken each day and Calculate and
report the mean and median total number of steps taken per day. Do these
values differ from the estimates from the first part of the assignment? What
is the impact of imputing missing data on the estimates of the total daily
number of steps?
```{r totalsteps_byday_noNA }
## We have already aggregated steps by day as totalsteps.
## Now we aggregate total steps by day without NAs
totalNoNAsteps <- aggregate(steps ~ date, activityNoNAs, sum)
```

```{r hist_comp_totalstepswith_without_NAs }
## Create a histogram of the aggregated steps by day data
par(mfrow=c(2,1))
hist(totalsteps$steps, breaks=10,
     main = paste("Histogram of Total Steps per Day"), xlab = "Total Steps",
     ylab ="Frequency of Days")

hist(totalNoNAsteps$steps, breaks=10,
     main = paste("Histogram of Total Steps per Day (NO NAs)"), xlab = "Total Steps",
     ylab ="Frequency of Days")
```

3. Calculate the median and mean of the aggregated steps by day data
```{r mean_median_comp_totalstepswith_without_NAs }
summary(totalsteps)
summary(totalNoNAsteps)

## Or calculate the median and mean individually
mean(totalsteps$steps)
median(totalsteps$steps)
mean(totalNoNAsteps$steps)
median(totalNoNAsteps$steps)
```

4. Find the mean of steps per interval and aggreate these into a new dataframe

```{r mean_steps_interval_with_withoutNAs}
stepinterval <- aggregate(steps ~ interval, activity, mean)
stepNoNAsinterval <- aggregate(steps ~ interval, activityNoNAs, mean)
```


## Are there differences in activity patterns between weekdays and weekends?

For this part the weekdays() function may be of some help here. Use the
dataset with the filled-in missing values for this part.

1. Create a new factor variable in the dataset with two levels – “weekday” and
“weekend” indicating whether a given date is a weekday or weekend day.

```{r weekday_weekend}
# convert date to a date() class variable
activityNoNAs$date <- as.Date(strptime(activityNoNAs$date, format="%Y-%m-%d"))

activityNoNAs$day <- weekdays(activityNoNAs$date)

for (i in 1:nrow(activityNoNAs)) {                                      
  if (activityNoNAs[i,]$day %in% c("Saturday","Sunday")) {            
    activityNoNAs[i,]$day<-"weekend"                                 
  }
  else{
    activityNoNAs[i,]$day<-"weekday"                                 
  }
}
```

2. Make a panel plot containing a time series plot of the 5-minute interval 
(x-axis) and the average number of steps taken, averaged
across all weekday days or weekend days (y-axis).

```{r plot_weekday_weekend}
activityNoNAsbyInt <- group_by(activityNoNAs,interval, day) 
activityNoNAsbyInt <- summarise(activityNoNAsbyInt, meansteps = mean(steps))

plot <- ggplot(activityNoNAsbyInt, aes(x = interval, y = meansteps)) + 
  geom_line() + facet_grid(day ~ .) + scale_x_continuous("Time of Day", breaks = seq(0, 2400, by = 200)) +
  scale_y_continuous("Mean number of steps", seq(0, 220, by = 30)) + ggtitle("Mean steps measured at 5 minute intervals")
plot
```

